
\chapter{Introduction and Background}
\label{ch:introduction} % for a text reference to this chapter: \ref{ch:introduction}

The \textit{Joint Spectral Radius} (JSR) was first introduced by G.-C. Rota and G. Strang in 1960 \citep{RotaStrang1960NoteJSR}. They described the JSR as the maximal exponential growth rate of a product of matrices taken from a finite set. Since its inception, the JSR has become a cornerstone in various mathematical and engineering disciplines due to its ability to encapsulate the asymptotic behavior of matrix products. 

The concept gained significant traction in the 1990s when researchers began exploring its theoretical properties and practical implications. Notable advancements include its application in wavelet theory, where it assists in the construction of refinable functions \citep{DaubechiesLagarias1992TwoScaleII}, as well as in control theory, where it is used to analyze the stability of switched linear systems \citep{Jungers2009JointSpectralRadius}, for which we provide a small example below. The computational challenges associated with determining the JSR have inspired the development of several algorithms, such as the invariant-polytope method \citep{GuglielmiProtasov2013ExactComputationJSR} and the finite-tree method \citep{MoellerReif2014TreeBasedJSR}.

Despite the progress, the JSR computation remains a challenging problem, particularly due to the exponential complexity of exploring all possible matrix products. This thesis seeks to contribute to this ongoing effort by leveraging the invariant-polytope algorithm and the finite-tree algorithm to create a hybrid methodology that mitigates their respective limitations.

\section*{Thesis Outline}
This thesis is structured as follows: Chapter~\ref{ch:introduction} provides background on the JSR and its basic properties. Chapters~\ref{ch:invariantPolytope} and~\ref{ch:finiteTree} present the key ideas behind the algorithms used to construct the proposed hybrid approach, including their theoretical foundations and algorithmic implementations. Chapter~\ref{ch:hybridApproach} discusses possible combinations of the former approaches, proposes the so-called Tree-flavored invariant-polytope algorithm, and provides proofs of its termination properties, which form the main results of this thesis. Chapter~\ref{ch:numericalTesting} presents numerical results to analyze the efficiency and applicability of the hybrid algorithm. Chapter~\ref{ch:thesisConclusion} concludes with insights and future directions.

\section{Notation and Preliminaries}
\label{sec:introPreliminaries}

In this thesis, the spectral radius of a matrix $A$ is denoted by $\rho(A)$. Unless stated otherwise, all matrices considered are assumed to have real entries, and all matrix norms are presumed to be submultiplicative. The analogous results for complex matrices follow similarly and may be reduced to the real case when necessary.

Let $\mathcal{A} = \{A_1, \dots, A_n\}$ denote a finite collection of matrices, each of size $d \times d$.

For any positive integer $k \in \mathbb{N}$, we define the set of positive multi-indices of length $k$ as

$$
    \mathcal{I}_k := \{1, \dots, n\}^k.
$$

Given a positive multi-index $I = [i_1, \dots, i_k] \in \mathcal{I}_k$, we define the associated matrix product by

$$
    A_I = A_{[i_1, \dots, i_k]} := A_{i_k} \cdots A_{i_1}.
$$
Note that the indices are in reversed order compared to the multiplications.

The set of all possible products of length $k$ formed from elements of $\mathcal{A}$ is then given by

$$
    \mathcal{A}^k := \left\{ A_I \mid I \in \mathcal{I}_k \right\}.
$$

For any scalar $\alpha \in \mathbb{R}$, we define the scaled matrix set by

$$
    \alpha \mathcal{A} := \left\{ \alpha A_1, \dots, \alpha A_n \right\}.
$$

\section{Motivation and Applications of the Joint Spectral Radius}
\label{sec:introMotivationJSR}

Let $\mathcal{A} = \{A_1, \cdots, A_n\}$ be a finite set of $d\times d$ matrices. 
Consider a discrete-time linear switched system of the form:
\begin{equation} \label{eq:switched_system}
x_{k+1} = A_{\sigma(k)} x_k, \quad x_0 \in \mathbb{R}^d,
\end{equation}
where $\sigma: \mathbb{N} \to \{1, \dots, n\}$ is a switching signal.

A key question is whether the system is \emph{uniformly asymptotically stable}, i.e., whether every trajectory $(x_k)$ converges to zero for all initial states and switching sequences:
\[
\lim_{k \to \infty} x_k = 0.
\]
This behavior is governed by a generalization of the spectral radius for a single matrix, the so-called \emph{joint spectral radius (JSR)} of a matrix set $\mathcal{A}$ \citep{RotaStrang1960NoteJSR}.
\begin{definition}\citep{RotaStrang1960NoteJSR}
    \label{def:jsrDefinition}
    For a finite set of matrices $\mathcal{A} = \{A_1, \dots, A_n\}$, the JSR is defined as:
    \begin{equation} \label{eq:jsr_def}
        \JSR(\mathcal{A}) := \lim_{k \to \infty} \max_{P \in \mathcal{A}^k}  \|P\|^{1/k}.
    \end{equation}
    where $\|\cdot\|$ denotes any matrix norm.
\end{definition}
This formulation captures the maximal asymptotic growth rate of matrix products.

\subsection*{Illustrative Example}

Consider the matrix set:
\[
\mathcal{A} = \left\{ A_1 = 
\begin{bmatrix}
0.5 & 1 \\
0 & 0.5
\end{bmatrix}, \quad
A_2 =
\begin{bmatrix}
0.3 & 0 \\
1 & 0.3
\end{bmatrix}
\right\}.
\]
Each matrix individually has spectral radius less than $1$. However, their product
\[
A_1 A_2 =
\begin{bmatrix}
0.5 & 1 \\
0 & 0.5
\end{bmatrix}
\begin{bmatrix}
0.3 & 0 \\
1 & 0.3
\end{bmatrix}
=
\begin{bmatrix}
1.3 & 0.3 \\
0.15 & 0.15
\end{bmatrix}
\]
has spectral radius of approximately
\[
\rho(A_1 A_2) \approx 1.2929 > 1.
\]
This example shows that switching can amplify the effects of individual matrices, leading to divergent behavior even if all single matrices are contractive. Thus, the JSR can exceed the spectral radius of any individual matrix in the set.

The following result motivates the importance of bounding or approximating $\JSR(\mathcal{A})$. It is known that:
\begin{itemize}
    \item The system~\eqref{eq:switched_system} is \emph{uniformly bounded} (i.e., $\sup_k \|x_k\| < \infty$ for all $x_0 \in \R^d$) if and only if $\JSR(\mathcal{A}) \leq 1$ and $\mathcal{A}$ is irreducible.
    \item It is \emph{uniformly asymptotically stable} if and only if $\JSR(\mathcal{A}) < 1$. 
\end{itemize}
Understanding the joint spectral radius (JSR) is therefore critical for analyzing the asymptotic behavior of switched linear systems, as illustrated by the example above, and motivates the development of reliable algorithms for determining or approximating it \citep{Jungers2009JointSpectralRadius}. Beyond control theory, the JSR plays a central role in several other areas. In approximation theory, it governs regularity questions for two-scale/refinement equations and wavelets \citep{DaubechiesLagarias1992TwoScaleII,Maesumi2000JSRHolderWavelets}. It is also used in the convergence and smoothness analysis of subdivision schemes \citep{GuglielmiManniVitale2011HermiteSubdivisionJSR}. In information theory, capacities of constrained codes can be studied through spectral characteristics of matrix families, leading to problems closely connected to JSR computation \citep{BlondelJungersProtasov2006CapacityCodes}. In discrete mathematics and combinatorics, JSR techniques appear in asymptotic counting problems such as partition-related functions and growth of structured matrix products \citep{Protasov2000PartitionFunctionAsymptotic,JungersProtasovBlondel2008IntegerMatrixGrowth,JungersProtasovBlondel2009OverlapFreeWordsSpectra}. These connections highlight the JSR as a unifying concept across operator theory, approximation theory, dynamical systems, information theory, and discrete mathematics, and further motivate efficient computational methods such as the hybrid algorithm based on works of \citep{MejstrikReif2025HybridApproachJSR}, which will be explored in this thesis.

\section{Fundamental Theoretical Results}
\label{sec:introTheoreticalBackground}

\begin{lemma}[Fekete's Lemma]\label{lem:feketeLemma}
Let $(a_n)_{n\ge 1}$ be a sequence of real, nonnegative numbers such that
\begin{equation}\label{eq:subadditive}
a_{m+n} \le a_m + a_n \qquad \text{for all } m,n\ge 1.
\end{equation}
Then the limit
\[
\lim_{n\to\infty}\frac{a_n}{n}
\]
exists and is equal to $ \inf_{n\ge 1} \left(\frac{a_n}{n} \right)$.
\end{lemma}

\begin{proof}
Set $b_n := a_n/n$ and let $\alpha := \inf_{n\ge 1} b_n$.
Fix $n\ge 1$ and write any $k\ge 1$ as $k = qn + r$ with integers $q\ge 0$ and
$0 \le r < n$. By subadditivity \eqref{eq:subadditive},
\[
a_k = a_{qn+r} \le q a_n + a_r.
\]
Dividing by $k$ gives
\[
\frac{a_k}{k} \le \frac{q a_n}{qn+r} + \frac{a_r}{qn+r}
\le \frac{a_n}{n} + \frac{a_r}{qn}.
\]
Taking $\limsup_{k\to\infty}$ (equivalently $q\to\infty$) yields
\[
\limsup_{k\to\infty}\frac{a_k}{k} \le \frac{a_n}{n}.
\]
Since $n$ was arbitrary, we obtain
\[
\limsup_{k\to\infty}\frac{a_k}{k} \le \inf_{n\ge 1}\frac{a_n}{n} = \alpha.
\]
On the other hand, by definition of infimum we always have
$\alpha \le a_k/k$ for all $k$, hence $\alpha \le \liminf_{k\to\infty} a_k/k$.
Therefore $\liminf = \limsup = \alpha$, so the limit exists and equals $\alpha$.
\end{proof}

\begin{theorem}\citep{RotaStrang1960NoteJSR,Jungers2009JointSpectralRadius}
    \label{thm:jsrWellDefined}
    The JSR is well-defined and independent of the choice of the matrix norm.
\end{theorem}

\begin{proof}
Let $\| \cdot \|_1$   and $ \| \cdot \|_2 $ be two matrix norms on $ \mathbb{R}^{n \times n} $. By equivalence of norms in finite-dimensional vector spaces, there exist constants $ c, C > 0 $ such that:
$$
c \|P\|_1 \leq \|P\|_2 \leq C \|P\|_1 \quad \forall P \in \mathbb{R}^{d \times d}
.$$
Therefore, we have

\begin{align*}
& \lim_{k \to \infty} \max_{P \in \mathcal{A}^k} \|P\|_{1}^{1/k} 
\le  \lim_{k \to \infty} {\left(\frac{1}{c}\right)}^{\frac{1}{k}} \max_{P \in \mathcal{A}^k} \|P\|_{2}^{1/k} \\
= & \lim_{k \to \infty} \max_{P \in \mathcal{A}^k} \|P\|_{2}^{1/k} 
\le  \lim_{k \to \infty} {C}^{\frac{1}{k}} \max_{P \in \mathcal{A}^k} \|P\|_{1}^{1/k}\\
= & \lim_{k \to \infty} \max_{P \in \mathcal{A}^k} \|P\|_{1}^{1/k},
\end{align*}
which makes the limit independent of the choice of the used matrix norm. 
The existence of this limit is a direct consequence of Fekete's Lemma.
\end{proof}

To facilitate the analysis of the algorithms and presentation of the main results, we first establish fundamental properties of the JSR.

\subsection*{Homogeneity of the Joint Spectral Radius}
\begin{proposition}\citep{Jungers2009JointSpectralRadius}
    \label{prop:jsrHomogeneity}
    The JSR is homogeneous, meaning for any set of matrices $\mathcal{A}$ and scalar $\alpha \in \R$ we have
    \begin{equation}
        \JSR(\alpha \mathcal{A}) = |\alpha| \JSR(\mathcal{A}).
    \end{equation}
\end{proposition}
\begin{proof}
    Let $\mathcal{A} = \{A_1, \dots, A_n\}$ and $\alpha \in \mathbb{R}$. Then:
    \begin{align*}
        \JSR(\alpha \mathcal{A}) & = \lim_{k \to \infty} \max_{A_{i} \in \mathcal{A}^k} \|\alpha A_{i_k} \cdots \alpha A_{i_1}\|^{1/k} \\
        & = |\alpha| \lim_{k \to \infty} \max_{A_{i} \in \mathcal{A}^k} \|A_{i_k} \cdots A_{i_1}\|^{1/k} \\
        & = |\alpha| JSR(\mathcal{A}) \\
        & \qedhere
    \end{align*}

\end{proof}

\subsection*{Berger--Wang Characterization and Three-Member Inequality}
\begin{proposition}\citep{BergerWang1992BoundedSemigroupsMatrices}
    \label{prop:bergerWangCharacterization}
    The $\JSR$ can equivalently be defined as
    $$
    \JSR(\mathcal{A}) = \limsup_{k \to \infty} \max_{P \in \mathcal{A}^k} \rho(P)^{1/k},
    $$
    if $\mathcal{A}$ is a finite (or compact) set.
\end{proposition}

\begin{proposition}\citep{Jungers2009JointSpectralRadius}\label{prop:threeMemberInequality}
Let $\|\cdot\|$ be a matrix norm. Then for every $k\in\mathbb{N}$,
\begin{equation}\label{eq:three_member}
\max_{P\in\mathcal{A}^k}\rho(P)^{1/k}\ \le\ \mathrm{JSR}(\mathcal{A})\ \le\ \max_{P\in\mathcal{A}^k}\|P\|^{1/k}.
\end{equation}
\end{proposition}

\begin{proof}
Fix $k\in\mathbb{N}$ and choose
\[
P_{\max}\in\operatorname*{arg\,max}_{P\in\mathcal{A}^k}\rho(P)^{1/k}.
\]
For the left-hand side we obtain, for any $r\in\mathbb{N}$,
\begin{equation}\label{eq:left_bound_three_member}
\begin{aligned}
\max_{P\in\mathcal{A}^k}\rho(P)^{1/k}
&= \rho(P_{\max})^{1/k} \\
&= \rho(P_{\max})^{r/(kr)} \\
&= \rho(P_{\max}^r)^{1/(kr)} \\
&\le \max_{P\in\mathcal{A}^{kr}}\rho(P)^{1/(kr)} \\
&\le \limsup_{r\to\infty}\max_{P\in\mathcal{A}^{kr}}\rho(P)^{1/(kr)} \\
&\le \limsup_{n\to\infty}\max_{P\in\mathcal{A}^n}\rho(P)^{1/n} \\
&= \mathrm{JSR}(\mathcal{A}).
\end{aligned}
\end{equation}

For the right-hand side, set
\[
M_n := \max_{P\in\mathcal{A}^n}\|P\|\qquad (n\in\mathbb{N}).
\]
By submultiplicativity of the norm, every $Q\in\mathcal{A}^{m+n}$ can be written
as $Q=RS$ with $R\in\mathcal{A}^m$ and $S\in\mathcal{A}^n$, hence
\begin{equation}\label{eq:Mn_submult}
\begin{aligned}
\|Q\|
&\le \|R\|\,\|S\|
\le M_m M_n,
\end{aligned}
\end{equation}
and taking maxima over $Q\in\mathcal{A}^{m+n}$ yields
\begin{equation}\label{eq:M_subadd}
\begin{aligned}
M_{m+n}\le M_m M_n
\qquad\text{for all } m,n\in\mathbb{N}.
\end{aligned}
\end{equation}

Define the real sequence $a_n := \log M_n$. Then from \eqref{eq:M_subadd},
\begin{equation}\label{eq:an_subadd}
\begin{aligned}
a_{m+n}
=\log M_{m+n}
\le \log(M_m M_n)
=\log M_m + \log M_n
=a_m+a_n,
\end{aligned}
\end{equation}
so $(a_n)_{n\ge 1}$ is subadditive. By Fekete's lemma, the limit
$\lim_{n\to\infty} a_n/n$ exists and equals $\inf_{n\ge 1} a_n/n$. In particular,
for the fixed $k$ we have
\begin{equation}\label{eq:fekete_step}
\begin{aligned}
\lim_{n\to\infty}\frac{a_n}{n}
=\inf_{n\ge 1}\frac{a_n}{n}
\le \frac{a_k}{k}.
\end{aligned}
\end{equation}
Exponentiating \eqref{eq:fekete_step} gives
\begin{equation}\label{eq:rhs_bound_three_member}
\begin{aligned}
\lim_{n\to\infty} M_n^{1/n}
&= \exp\!\Big(\lim_{n\to\infty}\frac{a_n}{n}\Big) \\
&\le \exp\!\Big(\frac{a_k}{k}\Big) \\
&= \exp\!\Big(\frac{\log M_k}{k}\Big) \\
&= M_k^{1/k}
= \max_{P\in\mathcal{A}^k}\|P\|^{1/k}.
\end{aligned}
\end{equation}

Since $\mathrm{JSR}(\mathcal{A})=\lim_{n\to\infty}\max_{P\in\mathcal{A}^n}\|P\|^{1/n}
=\lim_{n\to\infty}M_n^{1/n}$ by definition, the right-hand inequality in
\ref{eq:three_member} follows.
\end{proof}

This inequality underlies many computational methods, since it yields computable lower and upper bounds on $\mathrm{JSR}(\mathcal A)$. Moreover, as $k\to\infty$ the upper bound $\max_{P\in\mathcal A^k}\|P\|^{1/k}$ converges to $\mathrm{JSR}(\mathcal A)$, while the lower bound $\max_{P\in\mathcal A^k}\rho(P)^{1/k}$ approaches $\mathrm{JSR}(\mathcal A)$ in the sense of a $\limsup$.

\subsection*{Similarity Invariance and Reducibility}

\begin{proposition}\citep{Jungers2009JointSpectralRadius}
    \label{prop:jsrSimilarityInvariance}
    The $\JSR$ is invariant under similarity transformations, i.e., for any invertible matrix $T$ we have
    \begin{equation}
       \JSR(\mathcal{A}) = \JSR(T^{-1}\mathcal{A}T).
    \end{equation}
\end{proposition}

\begin{proof}
    This follows from the definition of the JSR and the properties of the spectral radius.
    \begin{align*}
    \JSR(T^{-1}\mathcal{A}T) & = \limsup_{k \to \infty} \max_{A_i \in \mathcal{A}^k} \rho(T^{-1}A_{i_k}T \cdot ... \cdot T^{-1}A_{i_1}T)^{1/k}\\
    & = \limsup_{k \to \infty} \max_{A_i \in \mathcal{A}^k} \rho(T^{-1}A_{i_k} \cdot ... \cdot A_{i_1}T)^{1/k}\\
    & = \limsup_{k \to \infty} \max_{A_i \in \mathcal{A}^k} \rho(A_{i_k} \cdot ... \cdot A_{i_1})^{1/k}\\
    & = \JSR(\mathcal{A})   
    \end{align*}
\end{proof}

\begin{definition}\citep{Jungers2009JointSpectralRadius}
    \label{def:matrixSetReducibility}
    A set of matrices is called \emph{(commonly) reducible} if there exists a nontrivial subspace of $\mathbb{R}^n$ that is invariant under all matrices in the set. This means there exists a change of basis that block-triangularizes all matrices in $\mathcal{A}$ at the same time. If $\mathcal{A}$ is not reducible, it is called \emph{irreducible}.
\end{definition}

\begin{proposition}\citep{Protasov1997GeneralizedSpectralRadiusGeometric}
    \label{prop:reducibleBlockDecomposition}
    If $\mathcal{A}$ is reducible, then 
    \begin{equation}
        \JSR(\mathcal{A}) = \max \{ \JSR(\mathcal{B}), \JSR(\mathcal{D}) \},
    \end{equation}
    where $\mathcal{B} = \{B_1,\dots, B_n\}$ and $\mathcal{D} = \{D_1,\dots, D_n\}$ are the blocks of the block-triangularized matrices
    \[
        T^{-1}A_jT = 
        \begin{bmatrix}
        B_j & C_j \\
        0   & D_j 
        \end{bmatrix}
    \]
    under some invertible matrix $T$.
\end{proposition}
\begin{example}
    \label{ex:reducibleMatricesBlockForm}
    \[
    A_1 = \begin{bmatrix} 2 & 1 \\ 1 & 2 \end{bmatrix}, \quad
    A_2 = \begin{bmatrix} 3 & 2 \\ 2 & 3 \end{bmatrix}
    \]

    These matrices are reducible because there exists a nontrivial change of basis that transforms them into block-triangular form. The eigenvectors common to both matrices are

    \[
    v_1 = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \quad
    v_2 = \begin{bmatrix} 1 \\ -1 \end{bmatrix}.
    \]

    Using the change of basis matrix

    \[
    T = \begin{bmatrix} 1 & 1 \\ 1 & -1 \end{bmatrix},
    \]

    we compute the similarity transformations

    \[
    T^{-1} A_1 T = \begin{bmatrix} 3 & 0 \\ 0 & 1 \end{bmatrix}, \quad
    T^{-1} A_2 T = \begin{bmatrix} 5 & 0 \\ 0 & 1 \end{bmatrix}.
    \]

    Since both matrices are upper triangular in this basis, they share a common invariant subspace, proving that they are reducible.
    Now, from Proposition~\ref{prop:reducibleBlockDecomposition}, we know that the $\JSR$ can be calculated as
    \[ 
    \JSR(\mathcal{A}) = \max \{ \JSR(\{3,5\}), \JSR(\{1,1\})\} = 5
    \]
\end{example}

This can be applied iteratively until the sets of blocks are all irreducible.
The problem is thereby split into similar problems of smaller dimension.
For the following considerations we can now assume $\mathcal{A}$ to be irreducible.

\subsection*{Finiteness Property and Spectrum-Maximizing Products}
\begin{definition}\citep{LagariasWang1995FinitenessConjecture}
    \label{def:finitenessProperty}
    A matrix set $\mathcal{A} = \{ A_1, \dots, A_n \}$ is said to possess the \emph{finiteness property} if there exists a finite sequence of matrices $A_{i_1}, \dots, A_{i_k}$ such that
    \begin{equation}
        \operatorname{JSR}(\mathcal{A}) = \|A_{i_k} \cdots A_{i_1}\|^{1/k}.
    \end{equation}
    Any such product is referred to as a \emph{spectrum-maximizing product} (s.m.p.).
\end{definition}
Although the finiteness property does not hold for all matrix sets, it is crucial for the effectiveness and termination of most algorithmic approaches.
This is shown, for example, in \citep{BlondelTheysVladimirov2003CounterexampleFiniteness}.

\subsection*{Computational Complexity of JSR Computation}

The computation of the joint spectral radius (JSR) is theoretically challenging due to several complexity results.

Approximating the JSR is \textbf{NP-hard} \citep{TsitsiklisBlondel1997HardToComputeJSR}, meaning no polynomial-time algorithm is expected unless $\text{P} = \text{NP}$. Moreover, for general sets of matrices, even deciding whether the JSR is strictly less than one is \textbf{undecidable} \citep{BlondelTsitsiklis2000NPHardMaxAlgebra}. Despite these theoretical limitations, many practical cases allow for efficient numerical estimation.

These complexity results highlight fundamental challenges. There are still applied settings where efficient approximation algorithms provide useful results.

\subsection*{Candidate Products and Generators}
Approximating the JSR, using the invariant-polytope or the finite-tree algorithms, requires identifying candidate products or \emph{generators} of the matrix set that contribute most significantly to the asymptotic growth rate. These generators are often derived through optimization techniques and their identification is a key step in computational algorithms.

\subsection*{Spectral Gap}
\begin{definition}\citep[Definition 1.9]{MejstrikReif2025HybridApproachJSR}
\label{def:spectralGap}
    A finite matrix family $\mathcal{A}$ is said to possess a \emph{spectral gap} if there exists a constant $\gamma \in [0,1)$ such that for every $k \in \mathbb{N}$ and every product $A_I \in \mathcal{A}^k$,
    \[
    \rho(A_I) = \JSR(\mathcal{A})^k
    \quad \text{or} \quad
    \rho(A_I) \le \gamma\,\JSR(\mathcal{A})^k.
    \]
    Equivalently, every non-s.m.p.\ product is uniformly separated from the maximal growth rate by the factor $\gamma < 1$.
\end{definition}

\begin{remark}\citep{MejstrikReif2025HybridApproachJSR,GuglielmiProtasov2024SpectralGapFamilyMatrices}
\label{rem:spectralGapHybrid}
    In the invariant-polytope setting, a spectral gap is part of the standard finite-termination assumptions together with uniqueness (up to cyclic permutations and powers) and simplicity of the leading eigenvalue of the s.m.p.\ candidate. 
    The hybrid/tree-based framework relaxes this: once preprocessing yields a correctly scaled candidate with $\rho(\Pi)=1$, proving existence of a suitable $V$-closed tree is sufficient to conclude $\JSR(\mathcal{A}) \le 1$, hence no explicit spectral-gap estimate is required for correctness.
    We also note that \citep{GuglielmiProtasov2024SpectralGapFamilyMatrices} studies a closely related notion of gap as separation between the first and second Lyapunov exponents for matrix families.
\end{remark}

\begin{example}
\label{ex:spectralGap}
    Consider
    \[
    A_1=\begin{bmatrix}2&0\\0&1\end{bmatrix},
    \qquad
    A_2=\begin{bmatrix}1&0\\0&2\end{bmatrix},
    \qquad
    \mathcal{A}=\{A_1,A_2\}.
    \]
    Since $\max\{\rho(A_1),\rho(A_2)\}=2$ and $\|A_1\|_\infty=\|A_2\|_\infty=2$, we get $\JSR(\mathcal{A})=2$.
    Any product $A_I \in \mathcal{A}^k$ has the form
    \[
    A_I = A_1^m A_2^{k-m}
        = \begin{bmatrix}2^m&0\\0&2^{k-m}\end{bmatrix},
    \]
    so
    \[
    \rho(A_I)=2^{\max\{m,k-m\}}.
    \]
    If $m=0$ or $m=k$, then $\rho(A_I)=2^k=\JSR(\mathcal{A})^k$ and $A_I$ is an s.m.p.\ product.
    If $1\le m\le k-1$, then
    \[
    \rho(A_I)\le 2^{k-1}=\tfrac12\,\JSR(\mathcal{A})^k.
    \]
    Hence Definition~\ref{def:spectralGap} holds with $\gamma=\tfrac12$.
\end{example}

\section{Preprocessing and Normalization}
\label{sec:introPreprocessing}
This thesis aims to address the challenge of computing the JSR by combining two existing algorithms that have demonstrated practical effectiveness in calculating the JSR for nontrivial sets of matrices. Both algorithms are based on the following simple concept:

We want to find the JSR of the finite set of matrices $\mathcal{A} = \{A_1, \cdots, A_n\}$.
\begin{enumerate}
    \item \textbf{Assumptions}: $\mathcal{A}$ is irreducible. 
    \item \textbf{Candidates}: Efficiently find products $P = A_{i_k} \cdots A_{i_1}$ of matrices from $\mathcal{A}$ that maximize the averaged spectral radius $\hat{\rho} := \rho(P)^\frac{1}{k}$.
    \item \textbf{Rescaling}: Transform $\mathcal{A} \to \tilde{\mathcal{A}}$ with $\tilde{A_i} := \frac{1}{\hat{\rho}} A_i$.
    \item \textbf{Verification}: Establish that JSR$(\tilde{\mathcal{A}}) = 1$ using the three-member inequality. By homogeneity, this is equivalent to JSR$(\mathcal{A}) = \hat{\rho}$.
\end{enumerate}

The considered algorithms differ only in step 4. The invariant-polytope algorithm tries to find a vector norm whose induced matrix norm already bounds the matrices from $\mathcal{A}$ sufficiently. The finite-tree algorithm, on the other hand, bounds products using a partitioning space that separates each product into terms bounded by 1 and a remainder term that does not grow fast enough to overcome the $k$-th root in the JSR definition (polynomial growth).
By integrating these algorithms into a hybrid approach, this work advances the computational tools available for JSR analysis by combining efficiency with a broad class of matrix sets for which the algorithm terminates.